A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-09 tex/sec5-synthesis.tex
[ tex ] Done with S5 for now

\begin{definition}\label{def:mode-correctness-args}
  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is defined as
  \begin{align}
    \label{eq:MC1} \MCas( \cdot ) & = \top \\
    \label{eq:MC3} \MCas\left(\biargvec, \chkbiarg[n]\right)
                                  & = \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land \MCas\left(\biargvec\right) \\
    \label{eq:MC2} \MCas\left(\biargvec, \synbiarg[n]\right) 
                                  & = \phantom{, A_n} \fv(\Delta) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land  \MCas\left(\biargvec\right)
  \end{align}
  where \eqref{eq:MC1} means an empty list is always mode-correct.
\end{definition}
This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\biargvec)$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
\begin{definition}\label{def:mode-correctness}
  An operation $\biop$ is \emph{mode-correct} if either
  \begin{enumerate}
    \item $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\biargvec)$ contains every inhabitant of $\Xi$; or
    \item $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\biargvec)$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
  \end{enumerate}
  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
\end{definition}
For a checking operation, an `input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each inhabitant of $\Xi$ during type synthesis.
On the other hand, for a synthesis operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
Also, the set of synthesis variables alone should include every type variable in $\Xi$ and particularly in $A_n$.
It is also easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct according to our definition.
\LT{perhaps we shall prove that mode-correctness is decidable}

@@

The interesting part of the proof is the case for the $\Rule{Op}$ rule, so we shall give the insight first before jumping into details.
Recall that a typing derivation for $\tmOpts$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with a \emph{partial} substitution $\rho_0$ defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere, as an `accumulating argument'.
Then, as we visit each synthesis argument $\synbiarg[i+1]$, we may be able to extend the domain of $\rho_i$ to include the synthesis variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and \emph{extending} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$.
By mode-correctness, the accumulating substitution~$\rho_i$ would be defined on enough synthesis variables so that each type synthesis and checking can be performed on $t_{i}$ with the context $\Gamma, \simsub{\Delta_{i+1}}{\rho_i}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
Then, if we go through every argument $t_i$ successfully, we will have a substitution $\rho_n$ from $\Xi$ to $\Type_{\Sigma}(\emptyset)$ defined on every variable $x : \Xi$ by mode-correctness.

To make the aforementioned argument holds, it is necessary to compare types and solve a unification problem.
Hence, we assume that the set $\Xi$ of variables has a decidable equality, rather than being an arbitrary set, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
To simplify our choice, we could simply confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$.
Given that these sets have a decidable equality and the arity of a type construct is finite, this would be a viable choice.
Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$ (see \Cref{sec:formalisation} for details).
However, for the sake of clarity in presentation, we keep using named variables and just assume that $\Xi$ has a decidable equality.}
The unification problem is as follows:
\begin{lemma}
  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
  \begin{enumerate}
    \item either there is a \emph{minimal extension} $\bar{\rho}$ of $\rho$ such that $\simsub{A}{\bar{\rho}} = B$, or 
    \item there is no extension $\sigma$ of $\rho$ such that $\simsub{A}{\sigma} = B$
  \end{enumerate}
  where by an extension $\sigma \geq \rho$ we mean that the domain of $\sigma$ contains the domain $\rho$ and $\sigma(x) = \rho(x)$ for every $x$ in the domain of $\rho$; by a minimal extension $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho}$ of $\rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma$ is an extension of $\rho$ with $P(\sigma)$.
\end{lemma}
This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
We are now ready for the decidability proof.

\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis-checking}}]
  We prove this statement by induction on the mode derivation\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
  The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness.
  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |- \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
  The first three cases follow essentially the same argument provided by \citet{Wadler2022}, but we still give the arguments here for the sake of completeness.
  The last case \Rule{Op} is new and has been discussed informally above.
  \begin{itemize}
    \item \SynRule{Var}: As $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}$, there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.

    \item \SynRule{Anno}: For $\erase{\Gamma} |-_{\Sigma, \Omega} (\isTerm{t} \annotate A)^{\syn}$, it is decidable whether $\Gamma |- \isTerm{t} \chk A$ by induction hypothesis.
      \begin{itemize}
        \item If $\Gamma |- \isTerm{t} \chk A$, then $\Gamma |- \isTerm{t \annotate A} \syn A$.
        \item If $\Gamma |/- t \chk A$ but $\Gamma |- \isTerm{t \annotate A} \syn$, then by inversion $\Gamma |- t \chk A$, leading to a contradiction.
      \end{itemize}

    \item \ChkRule{Sub}: If $\erase{\Gamma} |-_{\Sigma, \Omega}t^{\chk}$ by the rule $\ChkRule{Sub}$, then $\Gamma |-_{\Sigma, \Omega} t^{\syn}$ by inversion.
      We show that it is decidable for any $A$ whether $\Gamma |-_{\Sigma, \Omega} t \chk A$.
      By induction hypothesis, it is decidable whether $\Gamma |- t \syn B$ for some $B$:
      \begin{itemize}
        \item If $\Gamma |/- t \syn C$ for any type $C$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
        \item If $\Gamma |- t \syn B$ for some $B$, then either $A = B$ or $A \neq B$ by the decidable equality on $\Type_{\Sigma}(\Xi)$: 
          \begin{itemize}
            \item if $A = B$ then we have\/ $\Gamma |- t \chk A$;
            \item if $A \neq B$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn A$.
              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so we derive a contradiction.
          \end{itemize}
      \end{itemize}
    \item \Rule{Op}:
      Consider the mode derivation of\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \tmOpts^{\dir{d}}$.
%
%  \begin{claim}\label{lem:args-induction}
%    For the argument list $\overrightarrow{[\Delta_i] A_{i}^{\dir{d_i}}}$, \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$
%    \begin{enumerate}
%      \item either there is a minimal extension $\ext{\rho}$ of $\rho$ defined on $\synvar(\mathit{as})$ such that all of following judgements are derivable
%        \[
%          \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
%        \]
%
%      \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation. 
%    \end{enumerate}
%  \end{claim}
  \end{itemize}
\end{proof}

@@

%\begin{corollary}[Decidability of Bidirectional Type Synthesis] \label{thm:bidirectional-type-synthesis}
%  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
%  it is decidable whether there exists some $A$ such that
%  \[
%    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A.
%  \]
%\end{corollary}

>>>>

\begin{definition}\label{def:mode-correctness-args}
  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is defined as
  \begin{align}
    \label{eq:MC1} \MCas( \cdot ) & = \top \\
    \label{eq:MC3} \MCas\left(\biargvec, \chkbiarg[n]\right)
                                  & = \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land \MCas\left(\biargvec\right) \\
    \label{eq:MC2} \MCas\left(\biargvec, \synbiarg[n]\right) 
                                  & = \phantom{, A_n} \fv(\Delta) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land  \MCas\left(\biargvec\right)
  \end{align}
  where \eqref{eq:MC1} means an empty list is always mode-correct.
\end{definition}
This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\biargvec)$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
\begin{definition}\label{def:mode-correctness}
  An operation $\biop$ is \emph{mode-correct} if either
  \begin{enumerate}
    \item $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\biargvec)$ contains every inhabitant of $\Xi$; or
    \item $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\biargvec)$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
  \end{enumerate}
  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
\end{definition}
For a checking operation, an `input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each inhabitant of $\Xi$ during type synthesis.
On the other hand, for a synthesis operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
Also, the set of synthesis variables alone should include every type variable in $\Xi$ and particularly in $A_n$.
It is also easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct according to our definition.
\LT{perhaps we shall prove it decidable}

@@

The interesting part of the proof is the case for the $\Rule{Op}$ rule, so we shall give the insight first instead of jumping into the details.
Recall that a typing derivation for $\tmOpts$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with a \emph{partial} substitution $\rho_0$, as an `accumulating argument', defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
Then, as we visit each synthesis argument $\synbiarg[i+1]$, we may be able to extend the domain of $\rho_i$ to include the synthesis variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and \emph{extending} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$.
By mode-correctness, the accumulating substitution~$\rho_i$ would be defined on enough synthesis variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \simsub{\Delta_{i+1}}{\rho_i}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
Then, if we go through every argument $t_i$ successfully, we will have a substitution $\rho_n$ defined everywhere by mode-correctness.

\begin{remark}
To make the argument above sound, it is necessary to compare types and solve a unification problem.
Hence, we assume that the set\/ $\Xi$ of variables has a decidable equality, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
To simplify our choice, we could simply confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$, given that these sets have a decidable equality and the arity of a type construct is finite.
Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$ (see \Cref{sec:formalisation} for details).
For the sake of clarity in presentation, though, we keep using named variables and just assume that $\Xi$ has a decidable equality.}
\end{remark}
We need some auxiliary definitions for the notion of extension to state the unification problem:
\begin{definition}
  Let $\rho$ be a partial substitution.
  Then, by an \emph{extension}\/ $\sigma \geq \rho$ we mean that the domain of $\sigma$ contains the domain $\rho$ and\/ $\sigma(x) = \rho(x)$ for every\/ $x$ in the domain of $\rho$; by a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
\end{definition}
\begin{lemma}\label{lem:unify}
  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution\/ $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
  \begin{enumerate}
    \item either there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$, or 
    \item there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
  \end{enumerate}
\end{lemma}
This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
We are now ready for the decidability proof.

\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis-checking}}]
  We prove this statement by induction on the mode derivation\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
  The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness.
  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
  The first three cases follow essentially the same argument provided by \citet{Wadler2022}, but we still give the arguments here for the sake of completeness.
  The last case \Rule{Op} is new and has been discussed above.
  (For brevity we omit the subscript $(\Sigma, \Omega)$.)
  \begin{itemize}
    \item \SynRule{Var}: As $\erase{\Gamma} |- \isTerm{t}$, there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.

    \item \SynRule{Anno}: For $\erase{\Gamma} |- (\isTerm{t} \annotate A)^{\syn}$, it is decidable whether $\Gamma |- \isTerm{t} \chk A$ by induction hypothesis.
      \begin{itemize}
        \item If $\Gamma |- \isTerm{t} \chk A$, then $\Gamma |- \isTerm{t \annotate A} \syn A$.
        \item If $\Gamma |/- t \chk A$ but $\Gamma |- \isTerm{t \annotate A} \syn$, then by inversion $\Gamma |- t \chk A$, leading to a contradiction.
      \end{itemize}

    \item \ChkRule{Sub}: If $\erase{\Gamma} |-t^{\chk}$ by the rule $\ChkRule{Sub}$, then $\Gamma |- t^{\syn}$ by inversion.
      We show that it is decidable for any $A$ whether $\Gamma |- t \chk A$.
      By induction hypothesis, it is decidable whether $\Gamma |- t \syn B$ for some $B$:
      \begin{itemize}
        \item If $\Gamma |/- t \syn C$ for any $C$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
        \item If $\Gamma |- t \syn B$ for some $B$, then either $A = B$ or $A \neq B$ by the decidable equality on $\Type_{\Sigma}(\Xi)$: 
          \begin{itemize}
            \item if $A = B$ then we have\/ $\Gamma |- t \chk A$;
            \item if $A \neq B$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn A$.
              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so we derive a contradiction.
          \end{itemize}
      \end{itemize}
    \item \Rule{Op}:
      Consider the mode derivation of $\erase{\Gamma} |- \tmOpts^{\dir{d}}$.
      First we claim:
      \begin{claim}\label{lem:args-induction}
        For an argument list $\as = \biargvec$ and any partial substitution $\rho$ from $\Xi$ to $\emptyset$
        \begin{enumerate}
          \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that 
            \begin{equation} \label{eq:claim}
              \text{$\ext{\rho}$ is defined on $\synvar(\as)$ and\/} \quad
              \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |- \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
              \quad\text{for all $i$};
            \end{equation}
          \item or there is no extension $\sigma$ of $\rho$ such that \eqref{eq:claim} holds.
        \end{enumerate}
      \end{claim}
      Then, case analysis on $\dir{d}$:
      \begin{itemize}
        \item $\dir{d}$ is $\syn$: We apply our claim with the partial substitution $\rho_0$ defined nowhere.
          \begin{enumerate}
            \item If there is no $\sigma \geq \rho$ such that \eqref{eq:claim} holds but for some $A$ we have $\Gamma |- \tmOpts \syn A$, then by inversion we have a substitution $\rho\colon \Sub{\Xi}{\emptyset}$ such that ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}}$ for every $i$.
              Obviously, $\rho$ is an extension of $\rho_0$ such that ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}}$ for every $i$ and it contradicts the assumption that no such an extension exists.

            \item If there exists a minimal $\ext{\rho} \geq \rho_0$ defined on $\synvar(\biargs)$ such that \eqref{eq:claim} holds, then by mode-correctness we have $\ext{\rho}$ is total and
              \[
                \Gamma |- \tmOpts \syn \simsub{A_0}{\ext{\rho}}.
              \]
          \end{enumerate}

        \item $\dir{d}$ is $\chk$: Let $A$ be a type and apply \Cref{lem:unify} with $\rho_0$ defined nowhere.
          \begin{enumerate}
            \item If there is no $\sigma \geq \rho_0$ such that $\simsub{A_0}{\sigma} = A$ but $\Gamma |- \tmOpts \chk A$, then by inversion there is a substitution $\rho$ such that $A = \simsub{A_0}{\rho}$, thus a contradiction.
            \item If there is a minimal $\ext{\rho} \geq \rho_0$ such that $\simsub{A_0}{\ext{\rho}} = A$, then apply our claim with $\ext{\rho}$:
              \begin{enumerate}
                \item If there is no $\sigma \geq \ext{\rho}$ satisfying \eqref{eq:claim} but $\Gamma |- \tmOpts \chk A$, then by inversion there is $\gamma$ such that $\simsub{A_0}{\gamma} = A$ and also ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ for every $i$.
                  Given that $\ext{\rho} \geq \rho$ is minimal such that $\simsub{A_0}{\ext{\rho}} = A$, then $\gamma$ is an extension of $\ext{\rho}$ but by assumption no such an extension satisfying ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ exists, thus a contradiction.

                \item If there is a minimal $\ext{\ext{\rho}} \geq \ext{\rho}$ such that \eqref{eq:claim}, then by mode-correctness $\ext{\ext{\rho}}$ is total and
                  \[
                    \Gamma |- \tmOpts \chk \simsub{A_0}{\ext{\ext{\rho}}}
                  \]
                  where $\simsub{A_0}{\ext{\ext{\rho}}} = \simsub{A_0}{\ext{\rho}} = A$ since $\ext{\ext{\rho}}(x) = \ext{\rho}$ for every $x$ in the domain of $\ext{\rho}$.
              \end{enumerate}
          \end{enumerate}
      \end{itemize}
      \begin{claimproof}
        For a partial substitution $\rho$, we prove it by induction on the list $\biargvec$:
        \begin{enumerate}
          \item For the empty list, $\rho$ is the minimal extension of $\rho$ itself satisfying \eqref{eq:claim} trivially. 
          \item For $\biargvec, \biarg[m+1]$, by induction hypothesis on the list, we have two cases:
            \begin{enumerate}
              \item If there is no $\sigma \geq \rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m$ but a minimal $\gamma \geq \rho$ such that~\eqref{eq:claim} holds for all $1 \leq i \leq m + 1$, then we clearly have a contradiction.
              \item There is a minimal $\ext{\rho} \geq \rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m$.
                By case analysis on $\dir{d_{m+1}}$:
                \begin{itemize}
                  \item $\dir{d_{m+1}}$ is $\chk$: By mode-correctness, $\simsub{\Delta_{m+1}}{\ext{\rho}}$ and $\simsub{A_{m+1}}{\ext{\rho}}$ are defined and by the induction hypothesis it is decidable whether $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \chk \simsub{A_{m+1}}{\ext{\rho}}^{\dir{d_{m+1}}}$.
                    Clearly, if $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \chk \simsub{A_{m+1}}{\ext{\rho}}^{\dir{d_{m+1}}}$ then the desired statement is proved; otherwise we can also easily derive a contradiction.

                  \item $\dir{d_{m+1}}$ is $\syn$: By mode-correctness, $\simsub{\Delta_{m+1}}{\ext{\rho}}$ is defined and by the induction hypothesis, it is decidable that $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$.
                    \begin{enumerate}
                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |/- \isTerm{t_{m+1}} \syn A$ for any $A$ but there is $\gamma \geq \rho$ s.t.\ \eqref{eq:claim} holds for $1 \leq i \leq m+1$, then $\gamma \geq \ext{\rho}$.
                        Therefore $\simsub{\Delta_{m+1}}{\ext{\rho}} = \simsub{\Delta_{m+1}}{\gamma}$ and we derive a contradiction because $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$, then by \Cref{lem:unify} we have two cases: %we unify $A$ with $A_{m+1}$ extending $\ext{\rho}$.
                        \begin{itemize}
                          \item Suppose no $\sigma \geq \ext{\rho}$ such that $\simsub{A_{m+1}}{\sigma} = A$ but an extension $\gamma \geq \rho$ such that \eqref{eq:claim} holds for $1 \leq i \leq m + 1$. 
                            Then, $\gamma \geq \ext{\rho}$ by the minimality of $\ext{\rho}$ and thus
                            $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
                            However, by \Cref{thm:unique-syn}, the synthesised type $\simsub{A_{m+1}}{\gamma}$ must be unique, so $\gamma$ is an extension of $\ext{\rho}$ such that $\simsub{A_{m+1}}{\gamma} = A$, leading to a contradiction.
                          \item If there is a minimal $\ext{\ext{\rho}}$ such that $\simsub{A_{m+1}}{\ext{\ext{\rho}}} = A$, then it is now hard to show that $\ext{\ext{\rho}}$ is the minimal extension of $\rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m + 1$.
                        \end{itemize}
                    \end{enumerate}
                \end{itemize}
            \end{enumerate}
        \end{enumerate}
        \hfill{\footnotesize$\blacksquare$}
      \end{claimproof}
      We now complete the decidability proof.
  \end{itemize}
\end{proof}
While the proof presented above is complex, its formal counterpart essentially serves as two programs for checking and synthesis.
These programs either compute the typing derivation or provide a proof of contradiction.
Each case analysis simply branches based on the outcomes of bidirectional type synthesis and checking for each subterm, as well as the unification.
If a proof of contradiction is not of interest for implementation, the programs can be simplified by disregarding the cases that provide such contradiction proofs.
On the other hand, the contradiction proof contains the information as to why the construction of a derivation failed.
This can potentially be restructured further to provide informative error messages, assisting users in resolving issues with ill-typed terms, instead of merely answering a blatant `no'.

@@
